---
title: "Hands on Excercise 4"
author: "LIANG YAO"
date: '6 Dec 2023'
date-modified: "`r Sys.Date()`"
execute: 
  warning: false
  echo: true
  eval: true
format: 
  html: 
    code-fold: true
    code-summary: "Show the code"
---

## **1 Overview**

**Geographically weighted regression (GWR)** is a spatial statistical technique that takes non-stationary variables into consideration (e.g., climate; demographic factors; physical environment characteristics) and models the local relationships between these independent variables and an outcome of interest (also known as dependent variable). In this hands-on exercise, you will learn how to build [hedonic pricing](https://www.investopedia.com/terms/h/hedonicpricing.asp) models by using GWR methods. The dependent variable is the resale prices of condominium in 2015. The independent variables are divided into either structural and locational.

## **2 Preparing Data**

Two data sets will be used in this model building exercise, they are:

-   URA Master Plan subzone boundary in shapefile format (i.e. *MP14_SUBZONE_WEB_PL*)

-   condo_resale_2015 in csv format (i.e. *condo_resale_2015.csv*)

### **2.1 Get ready**

```{r}
pacman::p_load(olsrr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, tidyverse, gtsummary)
```

### **2.2 Geospatial Data**

Import the subzone data.

```{r}
mpsz <- st_read(dsn = "./data/geospatial",
                   layer = "MP14_SUBZONE_WEB_PL") %>% st_transform(crs = 3414)
```

Check the data.

```{r}
st_crs(mpsz)
```

Reveal the extent of mpsz.

```{r}
st_bbox(mpsz) #view extent
```

### **2.3 Aspatial Data**

Importing condo resale data.

```{r}

```

Wrangling data.
