---
title: "Hands on Exercise 2"
author: "LIANG YAO"
date: '22 Nov 2023'
date-modified: "`r Sys.Date()`"
execute: 
  warning: false
  echo: true
  eval: true
format: 
  html: 
    code-fold: true
    code-summary: "Show the code"
---

# **Spatial Weights and Applications**

## 1 Data Preparation

Two data sets will be used in this hands-on exercise, they are:

-   Hunan county boundary layer. This is a geospatial data set in ESRI shapefile format.

-   Hunan_2012.csv: This csv file contains selected Hunan's local development indicators in 2012.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

### 1.1 **Import shapefile into r environment**

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### 1.2 **Import csv file into r environment**

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package. The output is R dataframe class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 1.3 **Performing relational join**

The code chunk below will be used to update the attribute table of *hunan*'s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe. This is performed by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan,hunan2012)%>%
  select(1:4, 7, 15)
```

## 2 **Visualising Regional Development Indicator**

Now, we are going to prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 by using *qtm()* of **tmap** package.

```{r}
basemap <- tm_shape(hunan) +
  tm_polygons() +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC")
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

## 3 **Computing Contiguity Spatial Weights**

In this section, you will learn how to use [*poly2nb()*](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** package to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries. If you look at the documentation you will see that you can pass a "queen" argument that takes TRUE or FALSE as options. If you do not specify this argument the default is set to TRUE, that is, if you don't specify queen = FALSE this function will return a list of first order neighbours using the Queen criteria.

### 3.1 **Computing (QUEEN) contiguity based neighbours**

The code chunk below is used to compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, queen=TRUE)
summary(wm_q)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one heighbours.

For each polygon in our polygon object, *wm_q* lists all neighboring polygons. For example, to see the neighbors for the first polygon in the object, type:

```{r}
wm_q[[1]]
```

Polygon 1 has 5 neighbors. The numbers represent the polygon IDs as stored in hunan SpatialPolygonsDataFrame class.

We can retrive the county name of Polygon ID=1 by using the code chunk below:

```{r}
hunan$County[1]
```

The output reveals that Polygon ID=1 is Anxiang county.

To reveal the county names of the five neighboring polygons, the code chunk will be used:

```{r}
hunan$NAME_3[c(2,3,4,57,85)]
```

We can retrieve the GDPPC of these five countries by using the code chunk below.

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

The printed output above shows that the GDPPC of the five nearest neighbours based on Queen's method are 20981, 34592, 24473, 21311 and 22879 respectively.

You can display the complete weight matrix by using *str()*.

```{r}
str(wm_q) %>% head(10)
```

::: callout-warning
The output might cut across several pages. Save the trees if you are going to print out the report.
:::

### 3.2 **Creating (ROOK) contiguity based neighbours**

The code chunk below is used to compute Rook contiguity weight matrix.

```{r}
wm_r <- poly2nb(hunan, queen=FALSE)
summary(wm_r)
```

The summary report above shows that there are 88 area units in Hunan. The most connect area unit has 10 neighbours. There are two area units with only one heighbours.

### 3.3 **Visualising contiguity weights**

A connectivity graph takes a point and displays a line to each neighboring point. We are working with polygons at the moment, so we will need to get points in order to make our connectivity graphs. The most typically method for this will be polygon centroids. We will calculate these in the sf package before moving onto the graphs. Getting Latitude and Longitude of Polygon Centroids

We will need points to associate with each polygon before we can make our connectivity graph. It will be a little more complicated than just running st_centroid on the sf object: us.bound. We need the coordinates in a separate data frame for this to work. To do this we will use a mapping function. The mapping function applies a given function to each element of a vector and returns a vector of the same length. Our input vector will be the geometry column of us.bound. Our function will be st_centroid. We will be using map_dbl variation of map from the purrr package. For more documentation, check out map documentation

To get our longitude values we map the st_centroid function over the geometry column of us.bound and access the longitude value through double bracket notation \[\[\]\] and 1. This allows us to get only the longitude, which is the first value in each centroid.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

We do the same for latitude with one key difference. We access the second value per each centroid with \[\[2\]\].

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

Now that we have latitude and longitude, we use cbind to put longitude and latitude into the same object.

```{r}
coords <- cbind(longitude, latitude)
```

We check the first few observations to see if things are formatted correctly.

```{r}
head(coords)
```

#### 3.4.1 Plotting Queen contiguity based neighbours map

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "blue")
```

#### 3.4.2 Plotting Rook contiguity based neighbours map

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

#### 3.4.3 Plotting both Queen and Rook contiguity based neighbours maps

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col= "blue")
plot(hunan$geometry, border="lightgrey", main="Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

## 4 **Computing distance based neighbours**

In this section, you will learn how to derive distance-based weight matrices by using [*dnearneigh()*](https://r-spatial.github.io/spdep/reference/dnearneigh.html) of **spdep** package.

The function identifies neighbours of region points by Euclidean distance with a distance band with lower d1= and upper d2= bounds controlled by the bounds= argument. If unprojected coordinates are used and either specified in the coordinates object x or with x as a two column matrix and longlat=TRUE, great circle distances in **km** will be calculated assuming the WGS84 reference ellipsoid.

### 4.1 **Determine the cut-off distance**

Firstly, we need to determine the upper limit for distance band by using the steps below:

-   Return a matrix with the indices of points belonging to the set of the k nearest neighbours of each other by using [*knearneigh()*](https://r-spatial.github.io/spdep/reference/knearneigh.html) of **spdep**.

-   Convert the knn object returned by *knearneigh()* into a neighbours list of class nb with a list of integer vectors containing neighbour region number ids by using [*knn2nb()*](https://r-spatial.github.io/spdep/reference/knn2nb.html).

-   Return the length of neighbour relationship edges by using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**. The function returns in the units of the coordinates if the coordinates are projected, in km otherwise.

-   Remove the list structure of the returned object by using [**unlist()**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unlist).

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)         
```

The summary report shows that the largest first nearest neighbour distance is 61.79 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

### 4.2 **Computing fixed distance weight matrix**

Now, we will compute the distance weight matrix by using *dnearneigh()* as shown in the code chunk below.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

::: callout-tip
**Quiz:** What is the meaning of "Average number of links: 3.681818" shown above?

As I understand, this should be number of nonzero links over number of regions.
:::

Next, we will use *str()* to display the structure of the weight matrix is to combine [*table()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/table) and [*card()*](https://r-spatial.github.io/spdep/reference/card.html) of spdep.

```{r}
table(hunan$County, card(wm_d62))
```

To extract number of regions from the table:

```{r}
n_comp <- n.comp.nb(wm_d62)
table(n_comp$comp.id)
```

#### 4.2.1 Plotting fixed distance weight matrix

Next, we will plot the distance weight matrix by using the code chunk below.

```{r}
plot(hunan$geometry, border="lightgrey")
plot(wm_d62, coords, add=TRUE)
plot(k1, coords, add=TRUE, col="blue", length=0.08)
```

The blue lines show the links of 1st nearest neighbours and the black lines show the links of neighbours within the cut-off distance of 62km.

Alternatively, we can plot both of them next to each other by using the code chunk below.

```{r}
par(mfrow=c(1,2))
plot(hunan$geometry, border="lightgrey", main="1st nearest neighbours")
plot(k1, coords, add=TRUE, col="blue", length=0.08)
plot(hunan$geometry, border="lightgrey", main="Distance link")
plot(wm_d62, coords, add=TRUE, pch = 19, cex = 0.6)
```

### 4.3 **Computing adaptive distance weight matrix**

One of the characteristics of fixed distance weight matrix is that more densely settled areas (usually the urban areas) tend to have more neighbours and the less densely settled areas (usually the rural counties) tend to have lesser neighbours. Having many neighbours smoothes the neighbour relationship across more neighbours.

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry as shown in the code chunk below.

```{r}
knn6 <- knn2nb(knearneigh(coords, k=6))
knn6
```

::: callout-notice
Notice that each county has six neighbours, no less no more!
:::

#### 4.3.1 Plotting distance based neighbours

We can plot the weight matrix using the code chunk below.

```{r}
plot(hunan$geometry, border="lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "blue")
```

## 5 **Weights based on IDW**

In this section, you will learn how to derive a spatial weight matrix based on Inversed Distance method.

First, we will compute the distances between areas by using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**.

```{r}
dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(dist, function(x) 1/(x))
```

### 5.1 **Row-standardised weights matrix**

Next, we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight (style="W"). This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values. While this is the most intuitive way to summaries the neighbors' values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we'll stick with the style="W" option for simplicity's sake but note that other more robust options are available, notably style="B".

```{r}
rswm_q <- nb2listw(wm_q, style="W", zero.policy = TRUE)
rswm_q    
```

The zero.policy=TRUE option allows for lists of non-neighbors. This should be used with caution since the user may not be aware of missing neighbors in their dataset however, a zero.policy of FALSE would return an error.

To see the weight of the first polygon's eight neighbors type:

```{r}
rswm_q$weights[10]
```

Each neighbor is assigned a 0.125 of the total weight. This means that when R computes the average neighboring income values, each neighbor's income will be multiplied by 0.2 before being tallied.

Using the same method, we can also derive a row standardised distance weight matrix by using the code chunk below.

```{r}
rswm_ids <- nb2listw(wm_q, glist=ids, style="B", zero.policy=TRUE)
rswm_ids
```

```{r}
rswm_ids$weights[1]
```

```{r}
summary(unlist(rswm_ids$weights))
```

## 6 **Application of Spatial Weight Matrix**

In this section, you will learn how to create four different spatial lagged variables, they are:

-   spatial lag with row-standardized weights,

-   spatial lag as a sum of neighbouring values,

-   spatial window average, and

-   spatial window sum.

### 6.1 **Spatial lag with row-standardized weights**

Finally, we'll compute the average neighbor GDPPC value for each polygon. These values are often referred to as **spatially lagged values**.

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

Recalled in the previous section, we retrieved the GDPPC of these five countries by using the code chunk below.

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

::: callout-tip
Question: Can you see the meaning of Spatial lag with row-standardized weights now?
:::

We can append the spatially lag GDPPC values onto hunan sf data frame by using the code chunk below.

```{r}
lag.list <- list(hunan$NAME_3, lag.listw(rswm_q, hunan$GDPPC))
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan,lag.res)
```

The following table shows the average neighboring income values (stored in the Inc.lag object) for each county.

```{r}
head(hunan)
```

Next, we will plot both the GDPPC and spatial lag GDPPC for comparison using the code chunk below.

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp=1, ncol=2)
```

### 6.2 **Spatial lag as a sum of neighboring values**

We can calculate spatial lag as a sum of neighboring values by assigning binary weights. This requires us to go back to our neighbors list, then apply a function that will assign binary weights, then we use glist = in the nb2listw function to explicitly assign these weights.

We start by applying a function that will assign a value of 1 per each neighbor. This is done with lapply, which we have been using to manipulate the neighbors structure throughout the past notebooks. Basically it applies a function across each value in the neighbors structure.

```{r}
b_weights <- lapply(wm_q, function(x) 0*x + 1)
b_weights2 <- nb2listw(wm_q, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

With the proper weights assigned, we can use lag.listw to compute a lag variable from our weight and GDPPC.

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

::: callout-tip
Question: Can you understand the meaning of Spatial lag as a sum of neighboring values now?
:::

Next, we will append the *lag_sum GDPPC* field into `hunan` sf data frame by using the code chunk below.

```{r}
hunan <- left_join(hunan, lag.res)
```

Now, We can plot both the *GDPPC* and *Spatial Lag Sum GDPPC* for comparison using the code chunk below.

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp=1, ncol=2)
```

### 6.3 **Spatial window average**

The spatial window average uses row-standardized weights and includes the diagonal element. To do this in R, we need to go back to the neighbors structure and add the diagonal element before assigning weights.

To add the diagonal element to the neighbour list, we just need to use *include.self()* from **spdep**.

```{r}
wm_qs <- include.self(wm_q)
```

Notice that the Number of nonzero links, Percentage nonzero weights and Average number of links are 536, 6.921488 and 6.090909 respectively as compared to wm_q of 448, 5.785124 and 5.090909

Let us take a good look at the neighbour list of area \[1\] by using the code chunk below.

```{r}
wm_qs[[1]]
```

Notice that now \[1\] has six neighbours instead of five. Now we obtain weights with *nb2listw()*

```{r}
wm_qs <- nb2listw(wm_qs)
wm_qs
```

Again, we use *nb2listw()* and *glist()* to explicitly assign weight values.

Lastly, we just need to create the lag variable from our weight structure and GDPPC variable.

```{r}
lag_w_avg_gpdpc <- lag.listw(wm_qs, 
                             hunan$GDPPC)
lag_w_avg_gpdpc
```

Next, we will convert the lag variable listw object into a data.frame by using *as.data.frame()*.

```{r}
lag.list.wm_qs <- list(hunan$NAME_3, lag.listw(wm_qs, hunan$GDPPC))
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
```

Note: The third command line on the code chunk above renames the field names of *lag_wm_q1.res* object into *NAME_3* and *lag_window_avg GDPPC* respectively.

Next, the code chunk below will be used to append *lag_window_avg GDPPC* values onto *hunan* sf data.frame by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan, lag_wm_qs.res)
```

To compare the values of lag GDPPC and Spatial window average, `kable()` of Knitr package is used to prepare a table using the code chunk below. Let's check the first 10 rows.

```{r}
hunan %>%
  select("County", "lag GDPPC", "lag_window_avg GDPPC") %>%
  head(10) %>%
  kable()
```

Lastly, *qtm()* of **tmap** package is used to plot the lag_gdppc and w_ave_gdppc maps next to each other for quick comparison.

```{r}
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp=1, ncol=2)
```

::: callout-notice
For more effective comparison, it is advicible to use the core tmap mapping functions.
:::

### 6.4 **Spatial window sum**

The spatial window sum is the counter part of the window average, but without using row-standardized weights.

To add the diagonal element to the neighbour list, we just need to use *include.self()* from **spdep**.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

Next, we will assign binary weights to the neighbour structure that includes the diagonal element.

```{r}
b_weights <- lapply(wm_qs, function(x) 0*x + 1)
b_weights[1]
```

Notice that now \[1\] has six neighbours instead of five. Again, we use *nb2listw()* and *glist()* to explicitly assign weight values.

```{r}
b_weights2 <- nb2listw(wm_qs, 
                       glist = b_weights, 
                       style = "B")
b_weights2
```

With our new weight structure, we can compute the lag variable with *lag.listw()*.

```{r}
w_sum_gdppc <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
w_sum_gdppc
```

Next, we will convert the lag variable listw object into a data.frame by using *as.data.frame()*.

```{r}
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
```

Note: The second command line on the code chunk above renames the field names of *w_sum_gdppc.res* object into *NAME_3* and *w_sum GDPPC* respectively.

Next, the code chunk below will be used to append *w_sum GDPPC* values onto *hunan* sf data.frame by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan, w_sum_gdppc.res)
```

To compare the values of lag GDPPC and Spatial window average, `kable()` of Knitr package is used to prepare a table using the code chunk below.

```{r}
hunan %>%
  select("County", "lag_sum GDPPC", "w_sum GDPPC") %>%
  head(10) %>%
  kable()
```

Lastly, *qtm()* of **tmap** package is used to plot the lag_sum GDPPC and w_sum_gdppc maps next to each other for quick comparison.

```{r}
w_sum_gdppc <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(lag_sum_gdppc, w_sum_gdppc, asp=1, ncol=2)
```

# **Global Measures of Spatial Autocorrelation**

## 1 Importing Data into R

In this section, you will learn how to bring a geospatial data and its associated attribute table into R environment. The geospatial data is in ESRI shapefile format and the attribute table is in csv fomat.

### 1.1 **Import shapefile into r environment**

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### 1.2 **Import csv file into r environment**

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package. The output is R data frame class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

The code chunk below will be used to update the attribute table of *hunan*'s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe. This is performed by using *left_join()* of **dplyr** package.

```{r}
hunan <- left_join(hunan,hunan2012) %>%
  select(1:4, 7, 15)
```

### 1.3 **Visualising Regional Development Indicator**

Now, we are going to prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 by using *qtm()* of **tmap** package.

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## 2 **Global Spatial Autocorrelation**

In this section, you will learn how to compute global spatial autocorrelation statistics and to perform spatial complete randomness test for global spatial autocorrelation.

### 2.1 **Computing Contiguity Spatial Weights**

Before we can compute the global spatial autocorrelation statistics, we need to construct a spatial weights of the study area. The spatial weights is used to define the neighbourhood relationships between the geographical units (i.e. county) in the study area.

In the code chunk below, [*poly2nb()*](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** package is used to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries. If you look at the documentation you will see that you can pass a "queen" argument that takes TRUE or FALSE as options. If you do not specify this argument the default is set to TRUE, that is, if you don't specify queen = FALSE this function will return a list of first order neighbours using the Queen criteria.

More specifically, the code chunk below is used to compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

The summary report above shows that there are 88 area units in Hunan. The most connected area unit has 11 neighbours. There are two area units with only one neighbours.

### 2.2 **Row-standardised weights matrix**

Next, we need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight (style="W"). This is accomplished by assigning the fraction 1/(#ofneighbors) to each neighboring county then summing the weighted income values. While this is the most intuitive way to summaries the neighbors' values it has one drawback in that polygons along the edges of the study area will base their lagged values on fewer polygons thus potentially over- or under-estimating the true nature of the spatial autocorrelation in the data. For this example, we'll stick with the style="W" option for simplicity's sake but note that other more robust options are available, notably style="B".

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

The input of *nb2listw()* must be an object of class **nb**. The syntax of the function has two major arguments, namely style and zero.poly.

-   *style* can take values "W", "B", "C", "U", "minmax" and "S". B is the basic binary coding, W is row standardised (sums over all links to n), C is globally standardised (sums over all links to n), U is equal to C divided by the number of neighbours (sums over all links to unity), while S is the variance-stabilizing coding scheme proposed by Tiefelsdorf et al. 1999, p. 167-168 (sums over all links to n).

-   If *zero policy* is set to TRUE, weights vectors of zero length are inserted for regions without neighbour in the neighbours list. These will in turn generate lag values of zero, equivalent to the sum of products of the zero row t(rep(0, length=length(neighbours))) %\*% x, for arbitrary numerical vector x of length length(neighbours). The spatially lagged value of x for the zero-neighbour region will then be zero, which may (or may not) be a sensible choice.

### 2.3 **Moran's I**

In this section, you will learn how to perform Moran's I statistics testing by using [*moran.test()*](https://r-spatial.github.io/spdep/reference/moran.test.html) of **spdep**.

#### 2.3.1 **Maron's I test**

The code chunk below performs Moran's I statistical testing using [*moran.test()*](https://r-spatial.github.io/spdep/reference/moran.test.html) of **spdep**.

```{r}
moran.test(hunan$GDPPC, 
           listw=rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

::: callout-tip
Question: What statistical conclusion can you draw from the output above?

From ChatGPT:

Moran's I is a statistical measure used in spatial analysis to assess the spatial autocorrelation of data. Spatial autocorrelation refers to the relationship between the values of a variable in geographic space. In simpler terms, it explores whether nearby locations are similar or dissimilar in terms of a particular attribute or variable.

Moran's I statistic quantifies the degree of spatial clustering or dispersion in a dataset. It measures the correlation between the values of a variable at different locations. The values can range from -1 to 1, where:

-   Moran's I close to +1 indicates positive spatial autocorrelation, suggesting that similar values tend to cluster together in space. High values are indicative of a clustering pattern.

-   Moran's I close to -1 indicates negative spatial autocorrelation, suggesting dissimilar values tend to be close together in space. Low values are indicative of a dispersion or scattering pattern.

-   Moran's I close to 0 suggests no spatial autocorrelation, implying a random spatial pattern or no relationship between neighboring values.

As I understand, since the Moron I statistics standard deviate of 4.7351 indicates that the observed Moran's I value (0.300749970) is almost 4.7 standard deviations away from what would be expected under the null hypothesis. And p-value is very close to 0 suggesting to reject null hypothesis, indicating a significant positive spatial autocorrelation in our graph.
:::

#### 2.3.2 Computing Monte Carlo Moran's I

The code chunk below performs permutation test for Moran's I statistic by using [*moran.mc()*](https://r-spatial.github.io/spdep/reference/moran.mc.html) of **spdep**. A total of 1000 simulation will be performed.

```{r}
set.seed(1234)
bperm= moran.mc(hunan$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

::: callout-tip
Question: What statistical conclustion can you draw from the output above?

From ChaptGPT:

The Monte Carlo simulation of Moran's I is a method used to assess the statistical significance of Moran's I statistic by generating random spatial patterns through repeated simulations. This technique helps determine whether the observed spatial pattern in a dataset is significantly different from what would be expected under a null hypothesis of spatial randomness.

As I understand, this output shows a p-value close to 0, we should reject the null hypothesis again, and be confident in our conclusion that there is a statistically significant positive spatial autocorrelation in our graph.
:::

#### 2.3.3 Visualising Monte Carlo Moran's I

It is always a good practice for us the examine the simulated Moran's I test statistics in greater detail. This can be achieved by plotting the distribution of the statistical values as a histogram by using the code chunk below.

In the code chunk below [*hist()*](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/hist) and [*abline()*](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/abline) of R Graphics are used.

```{r}
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

**DIY: Using ggplot2 to plot.**

```{r}
res = as.data.frame(bperm$res)
ggplot(res, 
       aes(x= bperm$res))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  geom_vline(aes(xintercept = 0, color = "red"), show.legend = FALSE) +
  labs(title = "Distribution of Moran’s I Simulation",
      x = "Simulated Moran's I",
      y = "Frequency") +
  theme(panel.background = element_blank(), panel.grid = element_blank())
```

### 2.4 **Geary's**

In this section, you will learn how to perform Geary's c statistics testing by using appropriate functions of **spdep** package.

#### 2.4.1 Geary's C test

The code chunk below performs Geary's C test for spatial autocorrelation by using [*geary.test()*](https://r-spatial.github.io/spdep/reference/geary.test.html) of **spdep**.

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

::: callout-tip
Question: What statistical conclusion can you draw from the output above?

From ChatGPT:

Geary's C statistic measures spatial autocorrelation by comparing the differences between values at neighboring locations to the overall mean difference in the entire dataset. It can be used to detect both positive and negative spatial autocorrelation, indicating clustering or dispersion of attribute values across space.

Geary's C values range from 0 to infinity:

-   C\<1 suggests positive spatial autocorrelation (similar values are close together).

-   C\>1 suggests negative spatial autocorrelation (dissimilar values are close together).

-   C=1 implies no spatial autocorrelation (values are randomly distributed across space).

As I understand, since this test output 0.69 Geary's C statistic which is \<1, and very small p-value suggesting reject of null hypothesis, indicating there is autocorrelation in our graph and it's positive.
:::

#### 2.4.2 Computing Monte Carlo Geary's C

The code chunk below performs permutation test for Geary's C statistic by using [*geary.mc()*](https://r-spatial.github.io/spdep/reference/geary.mc.html) of **spdep**.

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

::: callout-tip
According to the very small p-value, we can safely reject the null hypothesis and trust our result from Geary C test.
:::

#### 2.4.3 Visualising the Monte Carlo Geary's C

```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

## 3 **Spatial Correlogram**

Spatial correlograms are great to examine patterns of spatial autocorrelation in your data or model residuals. They show how correlated are pairs of spatial observations when you increase the distance (lag) between them - they are plots of some index of autocorrelation (Moran's I or Geary's c) against distance.Although correlograms are not as fundamental as variograms (a keystone concept of geostatistics), they are very useful as an exploratory and descriptive tool. For this purpose they actually provide richer information than variograms.

### 3.1 **Compute Moran's I correlogram**

In the code chunk below, [*sp.correlogram()*](https://r-spatial.github.io/spdep/reference/sp.correlogram.html) of **spdep** package is used to compute a 6-lag spatial correlogram of GDPPC. The global spatial autocorrelation used in Moran's I. The **plot()** of base Graph is then used to plot the output.

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)
```

By plotting the output might not allow us to provide complete interpretation. This is because not all autocorrelation values are statistically significant. Hence, it is important for us to examine the full analysis report by printing out the analysis results as in the code chunk below.

```{r}
print(MI_corr)
```

::: callout-tip
Question:What statistical observation can you draw from the plot above?

Not all autocorrelation values are statistically significant.
:::

### 3.2 **Compute Geary's C correlogram and plot**

In the code chunk below, *sp.correlogram()* of **spdep** package is used to compute a 6-lag spatial correlogram of GDPPC. The global spatial autocorrelation used in Geary's C. The **plot()** of base Graph is then used to plot the output.

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)
```

Similar to the previous step, we will print out the analysis report by using the code chunk below.

```{r}
print(GC_corr)
```

# **Local Measures of Spatial Autocorrelation**

## 1 **Cluster and Outlier Analysis**

Local Indicators of Spatial Association or LISA are statistics that evaluate the existence of clusters in the spatial arrangement of a given variable. For instance if we are studying cancer rates among census tracts in a given city local clusters in the rates mean that there are areas that have higher or lower rates than is to be expected by chance alone; that is, the values occurring are above or below those of a random distribution in space.

In this section, you will learn how to apply appropriate Local Indicators for Spatial Association (LISA), especially local Moran'I to detect cluster and/or outlier from GDP per capita 2012 of Hunan Province, PRC.

## 2 **Computing local Moran's I**

To compute local Moran's I, the [*localmoran()*](https://r-spatial.github.io/spdep/reference/localmoran.html) function of **spdep** will be used. It computes *Ii* values, given a set of *zi* values and a listw object providing neighbour weighting information for the polygon associated with the zi values.

The code chunks below are used to compute local Moran's I of *GDPPC2012* at the county level.

```{r}
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)
```

*localmoran()* function returns a matrix of values whose columns are:

-   Ii: the local Moran's I statistics

-   E.Ii: the expectation of local moran statistic under the randomisation hypothesis

-   Var.Ii: the variance of local moran statistic under the randomisation hypothesis

-   Z.Ii:the standard deviate of local moran statistic

-   Pr(): the p-value of local moran statistic

The code chunk below list the content of the local Moran matrix derived by using [*printCoefmat()*](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/printCoefmat).

```{r}
printCoefmat(data.frame(
  localMI[fips,], 
  row.names=hunan$County[fips]),
  check.names=FALSE)
```

### 2.1 Mapping the local Moran's I

Before mapping the local Moran's I map, it is wise to append the local Moran's I dataframe (i.e. localMI) onto hunan SpatialPolygonDataFrame. The code chunks below can be used to perform the task. The out SpatialPolygonDataFrame is called *hunan.localMI*.

```{r}
hunan.localMI <- cbind(hunan,localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

### 2.2 Mapping local Moran's I values

Using choropleth mapping functions of **tmap** package, we can plot the local Moran's I values by using the code chinks below.

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty",
          palette = "RdBu",
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)
```

### 2.3 Mapping local Moran's I p-values

The choropleth shows there is evidence for both positive and negative Ii values. However, it is useful to consider the p-values for each of these values, as consider above.

The code chunks below produce a choropleth map of Moran's I p-values by using functions of **tmap** package.

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)
```

### 2.4 Mapping both local Moran's I values and p-values

For effective interpretation, it is better to plot both the local Moran's I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)
```

## 3 **Creating a LISA Cluster Map**

The LISA Cluster Map shows the significant locations color coded by type of spatial autocorrelation. The first step before we can generate the LISA cluster map is to plot the Moran scatterplot.

### 3.1 **Plotting Moran scatterplot**

The Moran scatterplot is an illustration of the relationship between the values of the chosen attribute at each location and the average value of the same attribute at neighboring locations.

The code chunk below plots the Moran scatterplot of GDPPC 2012 by using [*moran.plot()*](https://r-spatial.github.io/spdep/reference/moran.plot.html) of **spdep**.

```{r}
nci <- moran.plot(hunan$GDPPC, rswm_q,
                  labels=as.character(hunan$County), 
                  xlab="GDPPC 2012", 
                  ylab="Spatially Lag GDPPC 2012")
```

Notice that the plot is split in 4 quadrants. The top right corner belongs to areas that have high GDPPC and are surrounded by other areas that have the average level of GDPPC. This are the high-high locations in the lesson slide.

### 3.2 **Plotting Moran scatterplot with standardised variable**

First we will use [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) to centers and scales the variable. Here centering is done by subtracting the mean (omitting NAs) the corresponding columns, and scaling is done by dividing the (centered) variable by their standard deviations.

```{r}
hunan$Z.GDPPC <- scale(hunan$GDPPC) %>% 
  as.vector 
```

The [*as.vector()*](https://www.rdocumentation.org/packages/pbdDMAT/versions/0.5-1/topics/as.vector) added to the end is to make sure that the data type we get out of this is a vector, that map neatly into out dataframe.

Now, we are ready to plot the Moran scatterplot again by using the code chunk below.

```{r}
nci2 <- moran.plot(hunan$Z.GDPPC, rswm_q,
                   labels=as.character(hunan$County),
                   xlab="z-GDPPC 2012", 
                   ylab="Spatially Lag z-GDPPC 2012")
```

### 3.3 **Preparing LISA map classes**

The code chunks below show the steps to prepare a LISA cluster map.

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
```

Next, derives the spatially lagged variable of interest (i.e. GDPPC) and centers the spatially lagged variable around its mean.

```{r}
hunan$lag_GDPPC <- lag.listw(rswm_q, hunan$GDPPC)
DV <- hunan$lag_GDPPC - mean(hunan$lag_GDPPC)     
```

This is follow by centering the local Moran's around the mean.

```{r}
LM_I <- localMI[,1] - mean(localMI[,1])    
```

Next, we will set a statistical significance level for the local Moran.

```{r}
signif <- 0.05       
```

These four command lines define the low-low (1), low-high (2), high-low (3) and high-high (4) categories.

```{r}
quadrant[DV <0 & LM_I>0] <- 1
quadrant[DV >0 & LM_I<0] <- 2
quadrant[DV <0 & LM_I<0] <- 3  
quadrant[DV >0 & LM_I>0] <- 4      
```

Lastly, places non-significant Moran in the category 0.

```{r}
quadrant[localMI[,5]>signif] <- 0
```

### 3.4 **Plotting LISA map**

Now, we can build the LISA map by using the code chunks below.

```{r}
hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)
```

For effective interpretation, it is better to plot both the local Moran's I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
gdppc <- qtm(hunan, "GDPPC")

hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap <- tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

tmap_arrange(gdppc, LISAmap, 
             asp=1, ncol=2)
```

## 4 **Hot Spot and Cold Spot Area Analysis**

Beside detecting cluster and outliers, localised spatial statistics can be also used to detect hot spot and/or cold spot areas.

The term 'hot spot' has been used generically across disciplines to describe a region or value that is higher relative to its surroundings (Lepers et al 2005, Aben et al 2012, Isobe et al 2015).

### 4.1 **Getis and Ord's G-Statistics**

An alternative spatial statistics to detect spatial anomalies is the Getis and Ord's G-statistics (Getis and Ord, 1972; Ord and Getis, 1995). It looks at neighbours within a defined proximity to identify where either high or low values clutser spatially. Here, statistically significant hot-spots are recognised as areas of high values where other areas within a neighbourhood range also share high values too.

The analysis consists of three steps:

-   Deriving spatial weight matrix

-   Computing Gi statistics

-   Mapping Gi statistics

### 4.2 **Deriving distance-based weight matrix**

First, we need to define a new set of neighbours. Whist the spatial autocorrelation considered units which shared borders, for Getis-Ord we are defining neighbours based on distance.

There are two type of distance-based proximity matrix, they are:

-   fixed distance weight matrix; and

-   adaptive distance weight matrix.

#### 4.2.1 Deriving the centroid

We will need points to associate with each polygon before we can make our connectivity graph. It will be a little more complicated than just running *st_centroid()* on the sf object: **us.bound**. We need the coordinates in a separate data frame for this to work. To do this we will use a mapping function. The mapping function applies a given function to each element of a vector and returns a vector of the same length. Our input vector will be the geometry column of us.bound. Our function will be *st_centroid()*. We will be using map_dbl variation of map from the purrr package. For more documentation, check out map documentation

To get our longitude values we map the *st_centroid()* function over the geometry column of us.bound and access the longitude value through double bracket notation \[\[\]\] and 1. This allows us to get only the longitude, which is the first value in each centroid.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

We do the same for latitude with one key difference. We access the second value per each centroid with \[\[2\]\].

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

Now that we have latitude and longitude, we use cbind to put longitude and latitude into the same object.

```{r}
coords <- cbind(longitude, latitude)
```

#### 4.2.2 Determine the cut-off distance

Firstly, we need to determine the upper limit for distance band by using the steps below:

-   Return a matrix with the indices of points belonging to the set of the k nearest neighbours of each other by using [*knearneigh()*](https://r-spatial.github.io/spdep/reference/knearneigh.html) of **spdep**.

-   Convert the knn object returned by *knearneigh()* into a neighbours list of class nb with a list of integer vectors containing neighbour region number ids by using [*knn2nb()*](https://r-spatial.github.io/spdep/reference/knn2nb.html).

-   Return the length of neighbour relationship edges by using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**. The function returns in the units of the coordinates if the coordinates are projected, in km otherwise.

-   Remove the list structure of the returned object by using [**unlist()**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unlist).

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

The summary report shows that the largest first nearest neighbour distance is 61.79 km, so using this as the upper threshold gives certainty that all units will have at least one neighbour.

#### 4.2.3 Computing fixed distance weight matrix

Now, we will compute the distance weight matrix by using [*dnearneigh()*](https://r-spatial.github.io/spdep/reference/dnearneigh.html) as shown in the code chunk below.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62
```

Next, *nb2listw()* is used to convert the nb object into spatial weights object.

```{r}
wm62_lw <- nb2listw(wm_d62, style = 'B')
summary(wm62_lw)
```

### 4.3 **Computing adaptive distance weight matrix**

One of the characteristics of fixed distance weight matrix is that more densely settled areas (usually the urban areas) tend to have more neighbours and the less densely settled areas (usually the rural counties) tend to have lesser neighbours. Having many neighbours smoothes the neighbour relationship across more neighbours.

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry as shown in the code chunk below.

```{r}
knn <- knn2nb(knearneigh(coords, k=8))
knn
```

Next, *nb2listw()* is used to convert the nb object into spatial weights object.

```{r}
knn_lw <- nb2listw(knn, style = 'B')
summary(knn_lw)
```

## 5 **Computing Gi statistics**

### 5.1 **Gi statistics using fixed distance**

```{r}
fips <- order(hunan$County)
gi.fixed <- localG(hunan$GDPPC, wm62_lw)
gi.fixed
```

The output of localG() is a vector of G or Gstar values, with attributes "gstari" set to TRUE or FALSE, "call" set to the function call, and class "localG".

The Gi statistics is represented as a Z-score. Greater values represent a greater intensity of clustering and the direction (positive or negative) indicates high or low clusters.

Next, we will join the Gi values to their corresponding hunan sf data frame by using the code chunk below.

```{r}
hunan.gi <- cbind(hunan, as.matrix(gi.fixed)) %>%
  rename(gstat_fixed = as.matrix.gi.fixed.)
```

### 5.2 **Mapping Gi values with fixed distance weights**

The code chunk below shows the functions used to map the Gi values derived using fixed distance weight matrix.

```{r}
gdppc <- qtm(hunan, "GDPPC")

Gimap <-tm_shape(hunan.gi) +
  tm_fill(col = "gstat_fixed", 
          style = "pretty",
          palette="-RdBu",
          title = "local Gi") +
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, Gimap, asp=1, ncol=2)
```

### 2.3 **Gi statistics using adaptive distance**

The code chunk below are used to compute the Gi values for GDPPC2012 by using an adaptive distance weight matrix (i.e *knb_lw*).

```{r}
fips <- order(hunan$County)
gi.adaptive <- localG(hunan$GDPPC, knn_lw)
hunan.gi <- cbind(hunan, as.matrix(gi.adaptive)) %>%
  rename(gstat_adaptive = as.matrix.gi.adaptive.)
```

### 2.4 **Mapping Gi values with adaptive distance weights**

It is time for us to visualise the locations of hot spot and cold spot areas. The choropleth mapping functions of **tmap** package will be used to map the Gi values.

The code chunk below shows the functions used to map the Gi values derived using adaptive distance weight matrix.

```{r}
gdppc<- qtm(hunan, "GDPPC")

Gimap <- tm_shape(hunan.gi) + 
  tm_fill(col = "gstat_adaptive", 
          style = "pretty", 
          palette="-RdBu", 
          title = "local Gi") + 
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, 
             Gimap, 
             asp=1, 
             ncol=2)
```
