---
title: "Takehome_Ex01"
author: "LIANG YAO"
date: '25 Nov 2023'
date-modified: "`r Sys.Date()`"
execute: 
  warning: false
  echo: true
  eval: true
format: 
  html: 
    code-fold: true
    code-summary: "Show the code"
---

# **Geospatial Analytics for Public Good**

## **1 Setting the Scene**

As city-wide urban infrastructures such as buses, taxis, mass rapid transit, public utilities and roads become digital, the datasets obtained can be used as a framework for tracking movement patterns through space and time. This is particularly true with the recent trend of massive deployment of pervasive computing technologies such as GPS and RFID on the vehicles. For example, routes and ridership data were collected with the use of smart cards and Global Positioning System (GPS) devices available on the public buses. These massive movement data collected are likely to contain structure and patterns that provide useful information about characteristics of the measured phenomena. The identification, analysis and comparison of such patterns will provide greater insights on human movement and behaviours within a city. These understandings will potentially contribute to a better urban management and useful information for urban transport services providers both from the private and public sector to formulate informed decision to gain competitive advantage.

In real-world practices, the use of these massive locational aware data, however, tend to be confined to simple tracking and mapping with GIS applications. This is mainly due to a general lack of functions in conventional GIS which is capable of analysing and model spatial and spatio-temporal data effectively.

## **2 Objectives**

Exploratory Spatial Data Analysis (ESDA) hold tremendous potential to address complex problems facing society. In this study, you are tasked to apply appropriate Local Indicators of Spatial Association (GLISA) and Emerging Hot Spot Analysis (EHSA) to undercover the spatial and spatio-temporal mobility patterns of public bus passengers in Singapore.

## **3 The Data**

First of all, load needing packages.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr, sfdep, plotly)
```

### **3.1 Aspatial data**

For the purpose of this take-home exercise, *Passenger Volume by Origin Destination Bus Stops* downloaded from [LTA DataMall](https://www.waterpointdata.org/access-data/) will be used.

Import the passenger volume by origin destination bus stops data.

```{r}
odbus = read_csv("./data/aspatial/origin_destination_bus_202308.csv")  %>%
  mutate(ORIGIN_PT_CODE = as.factor(ORIGIN_PT_CODE),
         DESTINATION_PT_CODE = as.factor(DESTINATION_PT_CODE))
```

### **3.2 Geospatial data**

Two geospatial data will be used in this study, they are:

-   *Bus Stop Location* from LTA DataMall. It provides information about all the bus stops currently being serviced by buses, including the bus stop code (identifier) and location coordinates.

```{r}
busstop = st_read(dsn = "./data/geospatial/BusStopLocation_Jul2023",
                   layer = "BusStop")  %>% st_transform(crs = 3414)
```

-   *hexagon*, a [hexagon](https://desktop.arcgis.com/en/arcmap/latest/tools/spatial-statistics-toolbox/h-whyhexagons.htm) layer of 250m is provided to replace the relative coarse and irregular Master Plan 2019 Planning Sub-zone GIS data set of URA.

```{r}
hexagon = st_read(dsn = "./data/geospatial/hexagon",
                   layer = "hexagon")  %>% st_transform(crs = 3414)
```

## **4 The Task**

The specific tasks of this take-home exercise are as follows:

### **4.1 Geovisualisation and Analysis**

-   With reference to the time intervals provided in the table below, compute the passenger trips generated by origin at the hexagon level,

    | Peak hour period             | Bus tap on time |
    |------------------------------|-----------------|
    | Weekday morning peak         | 6am to 9am      |
    | Weekday afternoon peak       | 5pm to 8pm      |
    | Weekend/holiday morning peak | 11am to 2pm     |
    | Weekend/holiday evening peak | 4pm to 7pm      |

-   Display the geographical distribution of the passenger trips by using appropriate geovisualisation methods,

-   Describe the spatial patterns revealed by the geovisualisation (not more than 200 words per visual).

#### 4.1.1 Extracting Data

-   Weekday morning peak data.

    ```{r}
    weekday_6_9 <- odbus %>%
      filter(DAY_TYPE == "WEEKDAY") %>%
      filter(TIME_PER_HOUR >= 6 &
               TIME_PER_HOUR <= 9) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
      mutate(interval = "weekday_6_9")
    ```

-   Weekday afternoon peak data.

    ```{r}
    weekday_17_20 <- odbus %>%
      filter(DAY_TYPE == "WEEKDAY") %>%
      filter(TIME_PER_HOUR >= 17 &
               TIME_PER_HOUR <= 20) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
      mutate(interval = "weekday_17_20")
    ```

-   Weekend/holiday morning peak data.

    ```{r}
    weekend_11_14 <- odbus %>%
      filter(DAY_TYPE == "WEEKENDS/HOLIDAY") %>%
      filter(TIME_PER_HOUR >= 11 &
               TIME_PER_HOUR <= 14) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
      mutate(interval = "weekend_11_14")
    ```

-   Weekend/holiday evening peak data.

    ```{r}
    weekend_16_19 <- odbus %>%
      filter(DAY_TYPE == "WEEKENDS/HOLIDAY") %>%
      filter(TIME_PER_HOUR >= 16 &
               TIME_PER_HOUR <= 19) %>%
      group_by(ORIGIN_PT_CODE) %>%
      summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
      mutate(interval = "weekend_16_19")
    ```

-   Combine 4 time intervals data for further use.

    ```{r}
    peak_trips <- bind_rows(weekday_6_9, weekday_17_20, weekend_11_14, weekend_16_19)
    ```

#### 4.1.2 Wrangling geospatial data

First we need to combine bus stops and hexagon data.

```{r}
busstop_hexagon <- st_intersection(busstop, hexagon) %>%
  select(BUS_STOP_N,id) %>%
  st_drop_geometry()
```

Then combine passenger trip data with geospatial data.

```{r}
origin_trips <- left_join(peak_trips, busstop_hexagon,
            by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
  rename(ORIGIN_BS = ORIGIN_PT_CODE,
         ORIGIN_SZ = id) %>%
  group_by(ORIGIN_SZ) %>%
  summarise(TRIPS = sum(TRIPS))
```

Duplication check before continue:

```{r}
origin_trips %>%
  group_by_all() %>%
  filter(n()>1) %>%
  ungroup()
```

Re-join with hexagon data.

```{r}
peaktrip_hex <- left_join(hexagon, origin_trips, by = c("id" = "ORIGIN_SZ")) %>%
  filter(TRIPS>0)
```

#### 4.1.3 Visualizing passenger trips

4.1.3.1 Firstly, we may want check the geographical distribution of all bus trips to gather a full glimpse.

Below code chunk aims at wrangling the odbus data for visualization.

```{r}
odbus_by_ptcode <- odbus %>%
  select(ORIGIN_PT_CODE,TOTAL_TRIPS,DAY_TYPE) %>%
  group_by(ORIGIN_PT_CODE, DAY_TYPE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS)) 

bus_trips <- left_join(odbus_by_ptcode, busstop_hexagon,
            by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
  rename(ORIGIN_BS = ORIGIN_PT_CODE,
         ORIGIN_SZ = id) %>%
  group_by(ORIGIN_SZ) %>%
  summarise(TRIPS = sum(TRIPS))

bustrip_hex <- left_join(hexagon, bus_trips, by = c("id" = "ORIGIN_SZ")) %>%
  filter(TRIPS>0)
```

Now we can visualize the distribution of total bus trips.

```{r}
tmap_mode("view")
tm_shape(bustrip_hex) +
  tm_fill("TRIPS", 
          style = "quantile", 
          palette = "Blues",
          title = "Passenger trips",
          showNA = FALSE) +
  tm_layout(main.title = "Singapore Passenger Trips during Aug 2023",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_scale_bar() +
  tm_grid(alpha =0.2)
```

4.1.3.2 Then, let's check out the distribution of peak time trips of 4 time intervals in total.

```{r}
tmap_mode("view")
tm_shape(peaktrip_hex) +
  tm_fill("TRIPS", 
          style = "quantile", 
          palette = "Blues",
          title = "Passenger trips",
          showNA = FALSE) +
  tm_layout(main.title = "Peak Time Passenger Trips",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_scale_bar() +
  tm_grid(alpha =0.2)
```

::: callout-note
**Observation:**

There's no obvious difference between the distribution of total bus trips and peak time bus trips, so the peak time data we chose could be a meaningful representation of total trips.
:::

4.1.3.3 Then, to display the geographical distribution of 4 time intervals seperately for comparison:

-   Regenerate trip data with the "interval" column to indicate different time intervals.

    ```{r}
    peak_trips_interval <- left_join(peak_trips, busstop_hexagon,
                by = c("ORIGIN_PT_CODE" = "BUS_STOP_N")) %>%
      rename(ORIGIN_BS = ORIGIN_PT_CODE,
             ORIGIN_SZ = id) %>%
      group_by(ORIGIN_SZ, interval) %>%
      summarise(TRIPS = sum(TRIPS)) %>% 
      mutate(daily_trips = 
               ifelse(grepl("weekday",interval), ceiling(TRIPS/22), ceiling(TRIPS/9)))
    ```

    ::: callout-important
    Here I create a column "daily_trips" to find number of trips per day, since it's hard to compare the absolute number when we are differentiating weekday and weekend peak times intervals. For Aug 2023, there are 31 days in which 8 days are weekends and 1 day is National holiday.
    :::

    ```{r}
    peak_totaltrips_interval <- peak_trips_interval %>%
      select(ORIGIN_SZ,interval,TRIPS) %>%
      filter(TRIPS>0) %>%
      pivot_wider(names_from = interval, 
                  values_from = TRIPS, 
                  values_fill = NA)

    peak_dailytrips_interval <- peak_trips_interval %>%
      select(ORIGIN_SZ,interval,daily_trips) %>%
      filter(daily_trips>0) %>%
      pivot_wider(names_from = interval, 
                  values_from = daily_trips, 
                  values_fill = NA)
    ```

-   Join back with hexagon.

    ```{r}
    interval_totaltrip_hex <- left_join(hexagon, peak_totaltrips_interval,
                               by = c("id" = "ORIGIN_SZ")) 

    interval_dailytrip_hex <- left_join(hexagon, peak_dailytrips_interval,
                               by = c("id" = "ORIGIN_SZ")) 
    ```

-   Then we can draw graph for each time intervals.

    1\) Visualize total trip distribution for each time intervals.

    ```{r}
    tmap_mode("view")
    tm_shape(interval_totaltrip_hex)+ 
      tm_polygons(c("weekday_6_9","weekday_17_20","weekend_11_14","weekend_16_19"),
              style = "quantile", 
              palette = "Blues",
              title = "",
              colorNA = NULL,
              showNA = FALSE) + 
      tm_layout(panel.show = TRUE,
                panel.labels = c("Weekday 6-9am", "Weekday 5-8pm", "Weekend 11am-2pm", "Weekend 4-7pm")) 
    ```

    2\) Visualize daily trip distribution for each time intervals.

    ```{r}
    tmap_mode("view")
    tm_shape(interval_dailytrip_hex)+ 
      tm_polygons(c("weekday_6_9","weekday_17_20","weekend_11_14","weekend_16_19"),
              style = "quantile", 
              palette = "Blues",
              title = "",
              colorNA = NULL,
              showNA = FALSE) + 
      tm_layout(panel.show = TRUE,
                panel.labels = c("Weekday 6-9am", "Weekday 5-8pm", "Weekend 11am-2pm", "Weekend 4-7pm"))
    ```

::: callout-note
**Observation:**

1.  From the color scale we can find that the number of bus trips at night peak time intervals is larger than morning peak time intervals.

2.  Heavy trip hexagons are more disperse throughout the island during morning peak time intervals, and more concentrate in similar locations during night peak time intervals.
:::

### **4.2 Emerging Hot Spot Analysis(EHSA)**

With reference to the passenger trips by origin at the hexagon level for the four time intervals given above:

-   Perform Mann-Kendall Test by using the spatio-temporal local Gi\* values,

-   Prepared EHSA maps of the Gi\* values of the passenger trips by origin at the hexagon level. The maps should only display the significant (i.e. p-value \< 0.05).

-   With reference to the EHSA maps and data visualisation prepared, describe the spatial patterns reveled. (not more than 250 words per cluster).

#### 4.2.1 Perform Mann-Kendall Test

First we need to extract the passenger trips data during 4 peak time intervals from odbus.

```{r}
weekday_am_f <- odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 6 &
           TIME_PER_HOUR <= 9) %>% 
  group_by(TIME_PER_HOUR, ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
  rename('BUS_STOP_N'='ORIGIN_PT_CODE') %>%
  left_join(busstop_hexagon, by = 'BUS_STOP_N') %>% 
  group_by(TIME_PER_HOUR, id) %>%
  summarise(TRIPS = sum(TRIPS))

weekday_pm_f <-  odbus %>%
  filter(DAY_TYPE == "WEEKDAY") %>%
  filter(TIME_PER_HOUR >= 17 &
           TIME_PER_HOUR <= 20) %>% 
  group_by(TIME_PER_HOUR, ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
  rename('BUS_STOP_N'='ORIGIN_PT_CODE') %>%
  left_join(busstop_hexagon, by = 'BUS_STOP_N') %>% 
  group_by(TIME_PER_HOUR, id) %>%
  summarise(TRIPS = sum(TRIPS))

weekend_am_f <-  odbus %>%
  filter(DAY_TYPE == "WEEKENDS/HOLIDAY") %>%
  filter(TIME_PER_HOUR >= 11 &
           TIME_PER_HOUR <= 14) %>% 
  group_by(TIME_PER_HOUR, ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
  rename('BUS_STOP_N'='ORIGIN_PT_CODE') %>%
  left_join(busstop_hexagon, by = 'BUS_STOP_N') %>% 
  group_by(TIME_PER_HOUR, id) %>%
  summarise(TRIPS = sum(TRIPS))

weekend_pm_f <-  odbus %>%
  filter(DAY_TYPE == "WEEKENDS/HOLIDAY") %>%
  filter(TIME_PER_HOUR >= 16 &
           TIME_PER_HOUR <= 19) %>%
  group_by(TIME_PER_HOUR, ORIGIN_PT_CODE) %>%
  summarise(TRIPS = sum(TOTAL_TRIPS)) %>%
  rename('BUS_STOP_N'='ORIGIN_PT_CODE') %>%
  left_join(busstop_hexagon, by = 'BUS_STOP_N') %>% 
  group_by(TIME_PER_HOUR, id) %>%
  summarise(TRIPS = sum(TRIPS))
```

And then create an space-time cube for each time interval data.

```{r}
weekday_am_st <- spacetime(weekday_am_f, 
                     hexagon,
                      .loc_col = "id",
                      .time_col = "TIME_PER_HOUR")
```

Before continue, we better make sure the peak_st is indeed an space-time cube object.

```{r}
is_spacetime_cube(weekday_am_st)
```
