---
title: "Take Home Exercise 1"
author: "LIANG YAO"
date: '22 Nov 2023'
date-modified: "`r Sys.Date()`"
execute: 
  warning: false
  echo: true
  eval: true
format: 
  html: 
    code-fold: true
    code-summary: "Show the code"
---

# **Geospatial Analytics for Public Good**

## **Setting the Scene**

As city-wide urban infrastructures such as buses, taxis, mass rapid transit, public utilities and roads become digital, the datasets obtained can be used as a framework for tracking movement patterns through space and time. This is particularly true with the recent trend of massive deployment of pervasive computing technologies such as GPS and RFID on the vehicles. For example, routes and ridership data were collected with the use of smart cards and Global Positioning System (GPS) devices available on the public buses. These massive movement data collected are likely to contain structure and patterns that provide useful information about characteristics of the measured phenomena. The identification, analysis and comparison of such patterns will provide greater insights on human movement and behaviours within a city. These understandings will potentially contribute to a better urban management and useful information for urban transport services providers both from the private and public sector to formulate informed decision to gain competitive advantage.

In real-world practices, the use of these massive locational aware data, however, tend to be confined to simple tracking and mapping with GIS applications. This is mainly due to a general lack of functions in conventional GIS which is capable of analysing and model spatial and spatio-temporal data effectively.

## **Objectives**

Exploratory Spatial Data Analysis (ESDA) hold tremendous potential to address complex problems facing society. In this study, you are tasked to apply appropriate Local Indicators of Spatial Association (GLISA) and Emerging Hot Spot Analysis (EHSA) to undercover the spatial and spatio-temporal mobility patterns of public bus passengers in Singapore.

## **The Data**

### **Aspatial data**

For the purpose of this take-home exercise, *Passenger Volume by Origin Destination Bus Stops* downloaded from [LTA DataMall](https://www.waterpointdata.org/access-data/) will be used.

Import the passenger volume by origin destination bus stops data.

```{r}
#| eval: false
odbus = read_csv("./data/aspatial/origin_destination_bus_202308.csv")  %>%
  mutate(ORIGIN_PT_CODE = as.factor(ORIGIN_PT_CODE),
         DESTINATION_PT_CODE = as.factor(DESTINATION_PT_CODE))
```

### **Geospatial data**

Two geospatial data will be used in this study, they are:

-   *Bus Stop Location* from LTA DataMall. It provides information about all the bus stops currently being serviced by buses, including the bus stop code (identifier) and location coordinates.

```{r}
#| eval: false
busstop = st_read(dsn = "./data/geospatial/BusStopLocation_Jul2023",
                   layer = "BusStop")  %>% st_transform(crs = 3414)
```

-   *hexagon*, a [hexagon](https://desktop.arcgis.com/en/arcmap/latest/tools/spatial-statistics-toolbox/h-whyhexagons.htm) layer of 250m is provided to replace the relative coarse and irregular Master Plan 2019 Planning Sub-zone GIS data set of URA.

```{r}
#| eval: false
hexagon = st_read(dsn = "./data/geospatial/hexagon",
                   layer = "hexagon")  %>% st_transform(crs = 3414)
```

## **The Task**

The specific tasks of this take-home exercise are as follows:

### **Geovisualisation and Analysis**

-   With reference to the time intervals provided in the table below, compute the passenger trips generated by origin at the hexagon level,

    | Peak hour period             | Bus tap on time |
    |------------------------------|-----------------|
    | Weekday morning peak         | 6am to 9am      |
    | Weekday afternoon peak       | 5pm to 8pm      |
    | Weekend/holiday morning peak | 11am to 2pm     |
    | Weekend/holiday evening peak | 4pm to 7pm      |

-   Display the geographical distribution of the passenger trips by using appropriate geovisualisation methods,

-   Describe the spatial patterns revealed by the geovisualisation (not more than 200 words per visual).

### **Local Indicators of Spatial Association (LISA) Analysis**

-   Compute LISA of the passengers trips generate by origin at hexagon level.

-   Display the LISA maps of the passengers trips generate by origin at hexagon level. The maps should only display the significant (i.e. p-value \< 0.05)

-   With reference to the analysis results, draw statistical conclusions (not more than 200 words per visual).

### **Emerging Hot Spot Analysis(EHSA)**

With reference to the passenger trips by origin at the hexagon level for the four time intervals given above:

-   Perform Mann-Kendall Test by using the spatio-temporal local Gi\* values,

-   Prepared EHSA maps of the Gi\* values of the passenger trips by origin at the hexagon level. The maps should only display the significant (i.e. p-value \< 0.05).

-   With reference to the EHSA maps and data visualisation prepared, describe the spatial patterns reveled. (not more than 250 words per cluster).
